---
title: "Practical Machine Learning Course Project"
author: "Soma Chowdhury"
date: "2024-09-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Goal of this project is to predict the form of exercise done by participants, using data from activity recording devices such as Jawbone UP, Nike FuelBand and Fitbit.

In this project, we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. This is the “classe” variable in the training set. We train 4 models: **Decision Tree, Random Forest, Gradient Boosted Trees, Support Vector Machine** using k-folds cross validation on the training set. We then predict using a validation set randomly selected from the training csv data to obtain the **accuracy** and **out of sample error rate**. Based on those numbers, we decide on the best model, and use it to predict 20 cases using the test csv set.

## Background
Exercise enthusiasts quantify how much of a particular activity they do, but rarely quantify how well they do it.  
Our goal in this project is to use the data from accelerometers on the belt, forearm and dubmell of 6 participants.  
They were asked to lift the correctly and in correctly in 5 different ways.

## Data

The data in this project comes from  
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har

Training data set is available here  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Test data set is available here  
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

More information about the data is available here  
http://groupware.les.inf.puc-rio.br/har

## Loading Libraries and Data 

```{r results='hide', message=FALSE}
library(lattice)
library(ggplot2)
library(caret)
library(kernlab)
library(rattle)
library(corrplot)
set.seed(1234)
```

```{r}
trainData <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
testData <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")

dim(trainData)
dim(testData)
```

We see that their are 160 variables with 19622 observations in the training set and 20 in the test set.

## Cleaning the Data
Removing unnecessary variables. Starting with N/A values
```{r}
trainData <- trainData[,colMeans(is.na(trainData)) < .9] #removing mostly na columns
trainData <- trainData[,-c(1:7)] #removing metadata which is irrelevant to the outcome
```

Removing near zero variance variables.
```{r}
nvz <- nearZeroVar(trainData)
trainData <- trainData[,-nvz]
dim(trainData)
```
Now that we have finished removing the unnecessary variables, we can now split the training set into a **validation** and sub **training** set. The testing set "testData" will be left alone, and used for the final quiz test cases. 
```{r}
inTrain <- createDataPartition(y=trainData$classe, p=0.7, list=F)
train <- trainData[inTrain,]
valid <- trainData[-inTrain,]
```

## Creating and Testing the Models
Here we will test a few popular models including: **Decision Trees, Random Forest, Gradient Boosted Trees,** and **SVM.**

Set up control for training to use 3-fold cross validation. 
```{r}
control <- trainControl(method="cv", number=3, verboseIter=F)
```

## Decision Tree

**Model:** 

```{r, cache=TRUE}
mod_trees <- train(classe~., data=train, method="rpart", trControl = control, tuneLength = 5)
fancyRpartPlot(mod_trees$finalModel)
```

**Prediction:**

```{r}
pred_trees <- predict(mod_trees, valid)
cmtrees <- confusionMatrix(pred_trees, factor(valid$classe))
cmtrees
```

## Random Forest

```{r, cache=TRUE}
mod_rf <- train(classe~., data=train, method="rf", trControl = control, tuneLength = 5)

pred_rf <- predict(mod_rf, valid)
cmrf <- confusionMatrix(pred_rf, factor(valid$classe))
cmrf
```

## Gradient Boosted Trees

```{r, cache=TRUE}
mod_gbm <- train(classe~., data=train, method="gbm", trControl = control, tuneLength = 5, verbose = F)

pred_gbm <- predict(mod_gbm, valid)
cmgbm <- confusionMatrix(pred_gbm, factor(valid$classe))
cmgbm
```

## Support Vector Machine

```{r, cache=TRUE}
mod_svm <- train(classe~., data=train, method="svmLinear", trControl = control, tuneLength = 5, verbose = F)

pred_svm <- predict(mod_svm, valid)
cmsvm <- confusionMatrix(pred_svm, factor(valid$classe))
cmsvm
```

## Results (Accuracy & Out of Sample Error)

```{r, echo=FALSE}
models <- c("Tree", "RF", "GBM", "SVM")
accuracy <- round(c( cmtrees$overall[1], cmrf$overall[1], cmgbm$overall[1], cmsvm$overall[1]),3) #accuracy
oos_error <- 1 - accuracy #out of sample error

data.frame(accuracy = accuracy, oos_error = oos_error, row.names = models)
```

**The best model is the Random Forest model, with `r cmrf$overall[1]` accuracy and `r 1-cmrf$overall[1]` out of sample error rate. We find that to be a sufficient enough model to use for our test sets. ** 


# Predictions on Test Set

Running our test set to predict the classe (5 levels) outcome for 20 cases with the **Random Forest** model.
```{r}
pred <- predict(mod_rf, testData)
print(pred)
```

# Appendix

Github Repo: [Github](https://github.com/somachowdhury06/CourseraPracticalMachineLearning)

correlation matrix of variables in training set
```{r}
corrPlot <- cor(train[, -length(names(train))])
corrplot(corrPlot, method="color")
```

Plotting the models
```{r}
plot(mod_trees)
plot(mod_rf)
plot(mod_gbm)
```